Orientation Control System: Enhancing Aerial Maneuvers \\ for Quadruped Robots

### 2022-07-15 humanoids22roscia.tex
last modifications (Michele + Claudio comments) + rename file

\begin{abstract}%150-250 word abstract
For legged robots, aerial motions are the only option to overpass obstacles that cannot be circumvent with standard locomotion gaits. In these cases, the robot must perform a leap to either jump onto the obstacle or even fly over it. However, these movements represent a challenge because during the flight phase the center of mass cannot be controlled, and the robot orientation has limited controllability. This paper focuses on the latter issue and proposes an \gls{ocs} consisting of two rotating and actuated masses (flywheels or reaction wheels) to gain control authority on the robot orientation. Because of the conservation of angular momentum, their rotational velocity can be adjusted to steer the robot orientation even when there are no contacts with the ground. The axes of rotation of the flywheels are designed to be incident, leading to a compact orientation control system that is capable of controlling both roll and pitch angles, considering the different moment of inertia in the two directions. We tested the concept with simulations on the robot Solo12.
\end{abstract}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[width=.7\linewidth]{figures/solo_flywheels_pic.png}
	\caption{\small The proposed Orientation Control System for Solo12 has two flywheels with incident rotation axes. Each wheel is located in a 3D-printed shell and mounted on the trunk body.}
	\label{fig:solo12flywheels}
\end{figure}

Quadrupedal animals, like cats, can rearrange the tail and trunk to correct the orientation during a fall~\cite{kane1969dynamical}.
% Tail
Many works in the field of robotics used an additional link as a tail, like in \cite{chu2019null} and \cite{wenger2016frontal}.
This link rotates around an axis that does not pass through the robot \gls{com}: the distances between the axis of rotation and the \gls{com} of both trunk and tail result in a large effect on the total angular momentum even with a small tail mass.
However, the placement of the additional link makes the resulting robot asymmetric. Moreover, due to its limited range of motion, a tail can be used only for a single jump, not for a repeated sequence~\cite{johnson2012tail}.
% Legs reconfig
It is possible to obtain a similar result by creating repetitive circular motions with the feet, like in \cite{hoffman2021exploiting} and \cite{kurtz2021mini}. In the latter work, the authors proposed special heavy boots for Mini Cheetah and used a neural network to calculate online joint trajectories. However, this solution unnecessarily increases the inertia of the legs, complicating the locomotion problem, which can no longer rely on the massless leg assumption. %\annotation{trova un modo per dire che questa complicazione è data dalla massless assumption che non è più valida} 
% Gyros
Another option is to use a \gls{cmg}.
It consists of a wheel, spinning at a constant angular velocity inside two or three actuated gimbals.
Tilting the wheel's axis of rotation generates the gyroscopic torque.
This system is widespread in spacecrafts reorientation \cite{yoon2002spacecraft}, but less frequently exploited in robot locomotion, either wheeled \cite{brown1996single} or legged \cite{mikhalkov2021gyrubot}.
The \gls{cmg} presents interesting capabilities, but the presence of a pan-tilt unit to drive the gyroscope makes it impractical to mount it on a small, lightweight robot.

% Reaction wheel
Reaction wheels represent an additional option for controlling the robot orientation.
Changing the angular velocity of a rotating mass attached to the trunk generates a torque that can reorient the robot. This device has been first applied for spacecraft orientation \cite{oland2009reaction}, and later it has been only sporadically investigated in legged locomotion for controlling pitch orientation, both for bipeds \cite{Brown2016}, \cite{xiong2020sequential} and quadrupeds \cite{kolvenbach2019towards}, \cite{vasilopoulos2016quadruped}.
Compared with tails, reaction wheels do not have position limit, and since their rotation axes pass though the \gls{com}, its angular momentum results holonomic \cite{machairas2015quadruped}.
To get a fast response, it is necessary to have an abrupt change in the reaction wheel angular velocity (angular acceleration).
Using a brake avoids the employment of a motor able to deliver higher torques \cite{gajamohan2012cubli}, 
keeping the system compact: the motor slowly accelerates the wheel to a certain speed to store angular momentum and, when a reorientation is required, the break stops its spin.
Since the effect of the break is unidirectional, a limitation of this approach is given by the fact that it is possible to generate a rotation of the base only in the opposite direction of the reaction wheel angular velocity making them unsuitable for applications where \textit{continuous} controllability is required. On the other hand, direct-drive controlled flywheels enable to create accelerations in both directions,  and  to implement continuous control laws. The benefits of this kind of  \gls{ocs} are multiple: 1) flywheels can correct orientation errors during the flight (e.g. due to disturbances or tracking inaccuracies in the angular momentum achieved at the lift-off)   % and non-idealities).  
in a continuous manner. 2) They enable to track a time-varying reference (e.g. to have the robot land with a desired angular velocity (possibly zero) and orientation).
3) Even in presence of contacts, they can enhance the landing phase or stabilize dynamic gaits (e.g. trot), by significantly reducing trunk oscillations.  
4) The presence of these additional joints, whose only function is to control the orientation, gives the possibility to relieve the effort of the legs.
In more complex scenarios, like for a somersault, limbs and \gls{ocs} can operate in parallel to achieve a rotation angle larger than the one achievable only with legs (e.g., due to torque limitation).

\subsection{Proposed Approach and Contribution}
In this work, we present an \gls{ocs} based on two flywheels,  mounted on the trunk of the quadruped Solo12 \cite{grimminger2020open}, see Fig. \ref{fig:solo12flywheels}. 
The contributions of the paper are:
\begin{itemize}
	\item a novel design that enables to control the orientation of a legged robot during a jump in an effective way, while keeping the design simple. 
	In particular,  the axes of rotation of the flywheels are set to be incident, enabling \textit{continuous} controllability in both the directions (roll and pitch) while keeping the  device compact.
	\item simulations  with the quadruped Solo12 that demonstrate the effectivess of the proposed approach.
\end{itemize}

\subsection{Outline}
The reminder of this paper is organized as follow. In Section \ref{sec:background} the of conservation of the total angular momentum is recalled. In Section \ref{sec:ocs}, this principle is used in a simplified model for deriving the lower bound for the inertia of the \gls{ocs} and it is shown how to exploit it for simultaneously controlling the robot roll and pitch orientations. Section \ref{sec:simulations} presents simulation in different scenarios that proof the capability of our \gls{ocs} for rejecting disturbances and tracking angular references when there is no contact with the groung, and dumping base oscillation after the touch down. Conclusions and possible evolution of the work are reported in Section \ref{sec:conclusions}.
%

\section{Background}
\label{sec:background}
The starting point for any \gls{ocs} is Euler's equation. For any mechanical system, the time derivative of the angular momentum computed with the respect to a reference point $O$, fixed in an inertial frame,  equals the resultant torque with respect to the same reference point:
\begin{equation}
\dot{\bm{L}} = \sum_i \bm{M}_i
\end{equation}
When the resultant of the external moments applied to the system is zero, Euler’s equation simplifies to:
\begin{equation}
\dot{\bm{L}} = 0 \quad \Rightarrow \quad \bm{L}(t) = const,
\label{eq:cons_ang_mom}
\end{equation}
which is known as conservation of angular momentum.
Referring to legged robots, this condition occurs when the system is not in contact with the ground or other objects, e.g., during a fall or the flight phase of a jump.
In this case, it is possible to change the angular velocity of the base link changing the joint positions and velocities, as a result of the \textit{non-holonomy} of the total angular momentum \cite{Wieber16}: if the angular momentum of a certain body changes, the one of the others must change to maintain the total sum constant.

\section{Orientation Control System}
\label{sec:ocs}
In the following, we present a procedure for selecting the inertia of the flywheels, and we show how to exploit them for simultaneously controlling the robot roll and pitch orientation.
\subsection{Inertia Selection}
A preliminary analysis for the selection of the flywheels' inertia can be performed with the Elroy's Beanie model, depicted in Fig. \ref{fig:elroy-beanie}.
\begin{figure}
	\centering
	\includegraphics[width=.7\linewidth]{figures/elroys-beanie.eps}
	\caption{\small Schematic representation of the Elroy's Beanie model used for the preliminary analysis of the pitch motion.}
	\label{fig:elroy-beanie}
\end{figure} 
This consists of two rigid bodies connected through their \gls{com} with a revolute joint. One of the bodies represents the robot in its nominal configuration. The other models the \gls{ocs}, here represented by a single wheel for the sake of simplicity. The aim is to examine the rotational dynamics of the system as a whole. To simplify the analysis, in the following we will refer only to the effects on the pitch angle of the robot, keeping in mind that the same arguments apply also to the roll. 
Let us identify the moment of inertia of the robot in the nominal configuration as $I_r$ and the moment of inertia of the two flywheels as $I_f$, both referred to the system \gls{com}.
The angular momentum $L$ of this system can be written as
\begin{equation}\label{eq:ang_mom_elroy}
L(t) = \left(I_r + I_{f}\right)\dot{\theta}(t) + I_{f}\dot{\gamma}(t)
\end{equation}
where $\dot{\theta}$ and $\dot{\gamma}$ are respectively the robot pitch rate and the angular speed of the wheel.
We want to drive the robot to a desired pitch rate $\dot{\theta}_{des}$ acting on the wheel speed.
Without loss of generalities, we can assume the reaction wheel is stationary with the robot at the instant at which the reorientation maneuver starts ($\dot{\gamma}_0 = 0$), having a system angular momentum of $L_0 = \left(I_r + I_{f}\right)\dot{\theta}_0$.
Under the condition of conservation of the angular momentum, this quantity is constant over time and it is possible to estimate the lower bound for $I_{f}$ given a desired pitch rate and the maximum velocity of the flywheels $\dot{\gamma}_{max}$:
\begin{equation}
I_{f} \geq I_r \dfrac{ \Delta \dot{\theta} }{\Delta \dot{\theta}+\dot{\gamma}_{max}}.
\end{equation}
where $\Delta \dot{\theta} =  \left\lvert \dot{\theta}_0-\dot{\theta}_{des} \right\rvert$ is the base velocity variation.
Fig. \ref{fig:inertia_limits} reports the lower bound of $I_{f}$ given the desired base velocity variation and the maximum velocity of the actuator.
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{figures/lowerbound.eps}
	\caption{\small Minimum inertia of the \gls{ocs} (normalized by the robot inertia) necessary to achieve a base velocity variation $\Delta \dot{\theta}$, given the actuation bound $\dot \gamma^{max}$}
	\label{fig:inertia_limits}
\end{figure}

Once the desired inertia $I_f$ is chosen with the 2-D Elroy's Beanie model, it must be used to realize the 3-D \gls{ocs} (Fig. \ref{fig:flywheel}). 
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/flywheel_opaque.png}
	\caption{\small The final design of the reaction wheel, shown together with its inertia principal axes. Here, $h$ is the wheel thickness, $r$ is the inner radius of the wheel and $R$ is the outer one.}
	\label{fig:flywheel}
\end{figure}
We decided to realize the reaction wheels in the shape of hollow cylinders. This shape increases the inertia by locating the mass far away from the rotation axis. The inertia tensor expressed in its principal axes, $\bm{I}_f = \mathrm{diag}\left\{ I_{f, xx}, \, I_{f, yy}, \, I_{f, zz}  \right\}$, depends on the cylinders inner and outer radii, $r$ and $R$, its height $h$ and the material density $\rho$:
\begin{equation}
\begin{split}
I_{f, xx} = I_{f, yy} &= \frac{1}{12}\pi \rho h \left(3\left(R^4 - r^4\right) + h^2\left(R^2 - r^2\right)\right) \\
I_{f, zz} &= \frac{1}{2}\pi \rho h \left(R^4 - r^4\right)
\end{split}
\end{equation}
Notice that in the Elroy's Beanie model, there is a single body that models the complete \gls{ocs}. To meet the specification of controlling both the robot roll and pitch, it is possible to select one of two design options:
\begin{enumerate}
	\item use two couples of flywheels, each couple having the rotation axes aligned and orthogonal to the ones of the other couple \MF{metti foto con i due design uno di fianco all altro}; \ADP{Not clear why you would need 4 flywheels and not just 2. Also, I think it's not interesting to explain this design if then we don't use it.}
	\item use two flywheels, with incident axes of rotation.
\end{enumerate}
The first case would lead to a bulky \gls{ocs}, requiring brakes to achieve relevant angular momentum. Thus, the second option is considered. The inertia introduced in the Elroy's Beanie model has to be split between the two reaction wheels: $I_f = 2 \, I_{f, zz}$. \ADP{Not clear why the inertia should be split because you haven't discussed yet the fact that the flywheels won't be aligned with the axes. Move this to a later point.}
The parameters $R$ and $h$ can be set to have a compact \gls{ocs} and $\rho$ depends on the chosen material, that is stainless steel in our case. The inner radius $r$ can be adjusted to obtain the desired inertia. Spokes with negligible mass connect the wheel to the motor shaft. All the parameters are reported in Table \ref{tab:dim_fw}, together with the flywheels final inertia and mass.
\begin{table}
	\centering
	\caption{\small Size and dynamic parameters of a single reaction wheel}
	\begin{tabular}{ccc}
		Parameter & Value & Unit \\
		\hline
		\hline
		$r$ & $2.20\cdot 10^{-2}$ & $\mathrm{m}$ \\
		$R$ & $3.00\cdot 10^{-2}$ & $\mathrm{m}$ \\
		$h$   & $1.02\cdot 10^{-2}$ & $\mathrm{m}$ \\
		$\rho$   & $7.860 \cdot 10^{3}$ & $\mathrm{kg/m^3}$ \\
		$m$   & $1.02\cdot 10^{-1}$ & $\mathrm{kg}$ \\
		$I_{xx}, I_{yy}$   & $3.64 \cdot 10^{-5}$ & $\mathrm{kg \cdot m^2}$ \\
	$I_{zz}$   & $7.11 \cdot 10^{-5}$ & $ \mathrm{kg \cdot m^2}$ \\
	\end{tabular}
	\label{tab:dim_fw}
\end{table}

@@

The orientation of the axis of rotation of each flywheel influences their contribution to the total angular momentum. To meet the specification of controlling both the robot roll and pitch, the axes of rotation of the left and right wheel, identified in the base reference frame with the unit vectors $\hat{\bm{a}}_l$ and $\hat{\bm{a}}_r$, are set to be incident, laying on a plane parallel to the $xy-$plane of the base reference frame (see Fig. \ref{fig:axes}). 
To remove unnecessary complications, we design two symmetrical modules. We denote with $\alpha \leq \pi/2$ the non-negative incident angle between the wheels' axes of rotation and the robot lateral direction. The matrix 
\begin{equation}
\bm{C} = 
\left[ \begin{array}{cc}
\hat{\bm{a}}_l & \hat{\bm{a}}_r
\end{array} \right] = 
\left[ \begin{array}{cc}
\sin(\alpha) & -\sin(\alpha) \\
\cos(\alpha) & \cos(\alpha) \\
0 & 0
\end{array} \right]
\end{equation}
maps the reaction wheel moments into scalar torques $\bm{u} = \left[\begin{array}{cc} \tau_{fl} & \tau_{fr} \end{array} \right]^T$ about the flywheel axes $\hat{\bm{a}}_l$ and $\hat{\bm{a}}_r$ (expressed in the base frame).
As long as $\bm{C}$ is full column rank, it is possible to control both roll and pitch angles. If $\alpha = 0$, the roll results to be uncontrollable through the \gls{ocs}; otherwise, if $\alpha = \pi/2$, the pitch is uncontrollable. 
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/axes.eps}
	\caption{\small Representation of the flywheels rotation axes, seen from the top of the robot}
	\label{fig:axes}
\end{figure}
The angle is selected considering the ratio of the eigenvalues along the $x$ and $y$ directions of the ellipsoid of the robot inertia:
\begin{equation*}
\alpha^{*}=\tan^{-1}\left(\dfrac{I_{r, xx}}{I_{r, yy}}\right).
\end{equation*}
With these considerations, the angular momentum produced by the wheels, expressed in the robot base frame, is:
\begin{equation}
\begin{split}
\tensor[_b]{\bm{L}}{_f} &= \tensor[_b]{\bm{L}}{_{fl}} + \tensor[_b]{\bm{L}}{_{fr}} \\
&=I_{f, zz} \, \bm{\omega}_{fl} + I_{f, zz} \, \bm{\omega}_{fl}\\
&=I_{f, zz}
\left[\begin{array}{c}
\left(\omega_{fl} - \omega_{fr}\right)\sin(\alpha) \\ 
\left(\omega_{fl} +  \omega_{fr}\right)\cos(\alpha) \\
0
\end{array}\right]
\end{split}
\label{eq:bLf}
\end{equation}
in which $\bm{\omega}_{fl}= \omega_{fl} \, \hat{\bm{a}}_{l}$ and $\bm{\omega}_{fr}= \omega_{fr} \, \hat{\bm{a}}_{r}$ are the angular velocity vectors of the two wheels, and $\omega_{fl}$ and $\omega_{fr}$ are the speeds provided to each flywheel by its actuation system. The latter equation shows that the sum of the two angular speeds impacts on roll rotations, while their difference can be used to adjust the pitch, see Fig. \ref{fig:sum-diff}. Using the definition of $\bm{C}$, \eqref{eq:bLf} rewrites as
\begin{equation}
\begin{split}
\tensor[_b]{\bm{L}}{_f} & = I_{f,zz}
\left[
\begin{array}{cc}
\sin(\alpha) & \sin(\alpha) \\
\cos(\alpha) & -\cos(\alpha) \\
0 & 0
\end{array}
\right]
\left[
\begin{array}{cc}
\omega_{fl} \\
\omega_{fr}
\end{array}
\right] \\
&= I_{f,zz} \bm{C} \bm{\omega}_f.
\end{split}
\end{equation}
\begin{figure}
	\centering
	\includegraphics[width=.8\linewidth]{figures/sum_diff_mod.eps}
	\caption{\small Having incident rotation axes, the \gls{ocs} allows to control both the robot roll and pitch. Notice that the roll is influenced by the sum of the speeds of the flywheels, $\omega_x = \left(\omega_{fl} +  \omega_{fr}\right)\sin(\alpha)$. On the contrary, the pitch depends on their difference $\omega_y = \left(\omega_{fl} -  \omega_{fr}\right)\cos(\alpha)$.}
	\label{fig:sum-diff}
\end{figure}

>>>>

\begin{abstract}%150-250 word abstract
For legged robots, aerial motions are the only option to overpass obstacles that cannot be circumvent with standard locomotion gaits. In these cases, the robot must perform a leap to either jump onto the obstacle or fly over it. However, these movements represent a challenge because during the flight phase the \gls{com} cannot be controlled, and the robot orientation has limited controllability. This paper focuses on the latter issue and proposes an \gls{ocs} consisting of two rotating and actuated masses (flywheels or reaction wheels) to gain control authority on the robot orientation. Because of the conservation of angular momentum, their rotational velocity can be adjusted to steer the robot orientation even when there are no contacts with the ground. The axes of rotation of the flywheels are designed to be incident, leading to a compact orientation control system that is capable of controlling both roll and pitch angles, considering the different moment of inertia in the two directions. We tested the concept with simulations on the robot Solo12.
\end{abstract}

@@

\begin{figure}[t!]
	\centering
	\includegraphics[width=.7\linewidth]{figures/solo_fw_pic_compressed.png}
	\caption{\small The proposed Orientation Control System for 
		for the $2.5 \ \mathrm{kg}$ quadruped robot Solo12 consists of two $0.1 \ \mathrm{kg}$ flywheels with incident rotation axes. Each wheel is located in a 3D-printed shell and mounted on the trunk body.}
	\label{fig:solo12flywheels}
\end{figure}

Quadrupedal animals, like cats, can rearrange the tail and trunk to correct the orientation during a fall \cite{kane1969dynamical}.
% Tail
Many works in the field of robotics used an additional link as a tail, like in \cite{chu2019null} and \cite{wenger2016frontal}.
This link rotates around an axis that does not pass through the robot \gls{com}: the distances between the axis of rotation and the \gls{com} of both trunk and tail result in a large effect on the total angular momentum even with a small tail mass.
However, the placement of the additional link makes the resulting robot asymmetric. Moreover, due to its limited range of motion, a tail can be used only for a single jump, not for a repeated sequence~\cite{johnson2012tail}.
% Legs reconfig
It is possible to obtain a similar result by creating repetitive circular motions with the feet, like in \cite{hoffman2021exploiting} and \cite{kurtz2021mini}. In the latter work, the authors proposed special heavy boots for Mini Cheetah and used a neural network to calculate online joint trajectories. However, this solution unnecessarily increases the inertia of the legs, complicating the locomotion problem, which can no longer rely on the massless leg assumption. %\annotation{trova un modo per dire che questa complicazione è data dalla massless assumption che non è più valida} 
% Gyros
Another option is to use a \gls{cmg}.
It consists of a wheel, spinning at a constant angular velocity inside two or three actuated gimbals.
Tilting the wheel's axis of rotation generates the gyroscopic torque.
This system is widespread in spacecrafts reorientation \cite{yoon2002spacecraft}, but less frequently exploited in robot locomotion, either wheeled \cite{brown1996single} or legged \cite{mikhalkov2021gyrubot}.
The \gls{cmg} presents interesting capabilities, but the presence of a pan-tilt unit to drive the gyroscope makes it impractical to mount it on a small, lightweight robot.

% flywheel
flywheels represent an additional option for controlling the robot orientation.
Changing the angular velocity of a rotating mass attached to the trunk generates a torque that can reorient the robot. This device has been first applied for spacecraft orientation \cite{oland2009reaction}, and later it has been only sporadically investigated in legged locomotion for controlling pitch orientation, both for bipeds \cite{Brown2016}, \cite{xiong2020sequential} and quadrupeds \cite{kolvenbach2019towards}, \cite{vasilopoulos2016quadruped}.
Compared with tails, flywheels do not have position limits, and since their rotation axes pass through the \gls{com}, its angular momentum results holonomic \cite{machairas2015quadruped}.
To get a fast response, it is necessary to have an abrupt change in the flywheels angular velocity (angular acceleration).
Using a brake avoids the employment of a motor able to deliver higher torques \cite{gajamohan2012cubli}, 
keeping the system compact: the motor slowly accelerates the wheel to a certain speed to store angular momentum and, when a reorientation is required, the break stops its spin.
Since the effect of the break is unidirectional, a limitation of this approach is given by the fact that it is possible to generate a rotation of the base only in the opposite direction of the flywheel angular velocity making them unsuitable for applications where \textit{continuous} controllability is required. On the other hand, direct-drive controlled flywheels enable to create accelerations in both directions,  and  to implement continuous control laws. The benefits of this kind of  \gls{ocs} are multiple: 1) flywheels can correct orientation errors during the flight (e.g. due to disturbances or tracking inaccuracies in the angular momentum achieved at the lift-off)   % and non-idealities).  
in a continuous manner. 2) They enable to track a time-varying reference (e.g. to have the robot land with a desired angular velocity (possibly zero) and orientation).
3) Even in presence of contacts, they can enhance the landing phase or stabilize dynamic gaits (e.g. trot), by significantly reducing trunk oscillations.  
4) The presence of these additional joints, whose only function is to control the orientation, gives the possibility to relieve the effort of the legs.
In more complex scenarios, like for a somersault, limbs and \gls{ocs} can operate in parallel to achieve a rotation angle larger than the one achievable only with legs (e.g., due to torque limitation).

\subsection{Proposed Approach and Contribution}
In this work, we present a compact \gls{ocs} based on two flywheels, mounted on the trunk of the lightweight quadruped Solo12 \cite{grimminger2020open}, see Fig. \ref{fig:solo12flywheels}. 
The contributions of the paper are:
\begin{itemize}
	\item the design of a novel \gls{ocs} that enables to control the orientation of a legged robot during a jump in an effective way, while keeping the design simple. 
	In particular,  the axes of rotation of the flywheels are set to be incident, enabling \textit{continuous} controllability in both the directions (roll and pitch) while keeping the  device compact.
	\item simulations  with the quadruped Solo12 that demonstrate the effectiveness of the proposed approach.
\end{itemize}

\subsection{Outline}
The reminder of this paper is organized as follow. In Section \ref{sec:background}, the law of conservation of the total angular momentum is recalled. In Section \ref{sec:ocs}, the design principles are presented together with the strategy for simultaneously control the robot roll and pitch orientations. Section \ref{sec:simulations} presents simulation in different scenarios that demonstrate the the capability of our \gls{ocs} to reject disturbances and to track angular references when there is no contact with the ground, and damping of base oscillation after the touch-down. Conclusions and a possible evolution of the work ar reported in Section \ref{sec:conclusions}.
%

\section{Background}
\label{sec:background}
The starting point for any \gls{ocs} is the Euler's equation. For any mechanical system, the time derivative of the angular momentum $\bm{L}$ computed with the respect to a reference point $O$, fixed in an inertial frame, equals the sum of the moments $\bm{M}_i$ applied to the system with respect to the same reference point:
\begin{equation}
\dot{\bm{L}} = \sum_i \bm{M}_i
\end{equation}
When the resultant of the external moments applied to the system is zero, the Euler's equation simplifies to:
\begin{equation}
\dot{\bm{L}} = 0 \quad \Rightarrow \quad \bm{L}(t) = const,
\label{eq:cons_ang_mom}
\end{equation}
which is known as conservation of angular momentum.
Referring to legged robots, this condition occurs when the system is not in contact with the ground or other objects, e.g., during a fall or the flight phase of a jump.
In this case, it is possible to change the angular velocity of the base link changing the joint positions and velocities, as a result of the \textit{non-holonomy} of the total angular momentum \cite{Wieber16}: if the angular momentum of a certain body changes, the one of the others must change to maintain the total sum constant.

\section{Orientation Control System}
\label{sec:ocs}
In this section, we present a procedure to select the inertia of the flywheels, and we show how to exploit them for simultaneously controlling the robot roll and pitch orientation.
\subsection{Inertia Selection}
A preliminary analysis for the selection of the flywheels' inertia can be performed with the Elroy's Beanie model, depicted in Fig. \ref{fig:elroy-beanie}.
\begin{figure}
	\centering
	\includegraphics[width=.7\linewidth]{figures/elroys-beanie.eps}
	\caption{\small Schematic representation of the Elroy's Beanie model used for the preliminary analysis of the pitch motion.}
	\label{fig:elroy-beanie}
\end{figure} 
This consists of two rigid bodies connected through their \gls{com} with a revolute joint. One of the bodies represents the robot in its nominal configuration. The other models the \gls{ocs}, here represented by a single wheel for the sake of simplicity. The aim is to examine the rotational dynamics of the system as a whole. To simplify the analysis, in the following we will refer only to the effects on the pitch angle of the robot, keeping in mind that the same arguments apply also to the roll. 
Let us identify the moment of inertia of the robot in the nominal configuration as $I_r$ and the moment of inertia of the two flywheels as $I_f$, both referred to the system \gls{com}.
The angular momentum $L$ of this system can be written as
\begin{equation}\label{eq:ang_mom_elroy}
L(t) = \left(I_r + I_{f}\right)\dot{\theta}(t) + I_{f}\dot{\gamma}(t)
\end{equation}
where $\dot{\theta}$ and $\dot{\gamma}$ are respectively the robot pitch rate and the angular speed of the wheel.
We want to drive the robot to a desired pitch rate $\dot{\theta}_{des}$ by acting on the wheel speed.
Without loss of generality, we can assume the flywheel is stationary with the robot at the instant at which the reorientation maneuver starts ($\dot{\gamma}_0 = 0$), having a system angular momentum of $L_0 = \left(I_r + I_{f}\right)\dot{\theta}_0$.
Under the condition of conservation of the angular momentum, this quantity is constant over time and it is possible to estimate the lower bound for $I_{f}$ given a desired pitch rate and the maximum velocity of the flywheels $\dot{\gamma}_{max}$:
\begin{equation}
I_{f} \geq I_r \dfrac{ \Delta \dot{\theta} }{\Delta \dot{\theta}+\dot{\gamma}_{max}}.
\end{equation}
where $\Delta \dot{\theta} =  \left\lvert \dot{\theta}_0-\dot{\theta}_{des} \right\rvert$ is the base velocity variation.
Fig. \ref{fig:inertia_limits} reports the lower bound of $I_{f}$ given the desired base velocity variation and the maximum velocity of the actuator.
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{figures/lowerbound.eps}
	\caption{\small Minimum inertia of the \gls{ocs} (normalized by the robot inertia) necessary to achieve a base velocity variation $\Delta \dot{\theta}$, given the actuation bound $\dot \gamma^{max}$.}
	\label{fig:inertia_limits}
\end{figure}

@@

The orientation of the axis of rotation of each flywheel influences their contribution to the total angular momentum. To meet the specification of controlling both the robot roll and pitch, the axes of rotation of the left and right wheel, identified in the base reference frame with the unit vectors $\hat{\bm{a}}_l$ and $\hat{\bm{a}}_r$, are set to be incident, laying on a plane parallel to the $xy-$plane of the base reference frame (see Fig. \ref{fig:axes}). 
To remove unnecessary complications, we designed two identical modules, to be mounted on the trunk of Solo12, symmetrically with the respect of the robot sagittal plane. We denote with $\alpha \leq \pi/2$ the non-negative incident angle between the wheels' axes of rotation and the robot lateral direction. The matrix 
\begin{equation}
\bm{C} = 
\left[ \begin{array}{cc}
\hat{\bm{a}}_l & \hat{\bm{a}}_r
\end{array} \right] = 
\left[ \begin{array}{cc}
\sin(\alpha) & -\sin(\alpha) \\
\cos(\alpha) & \cos(\alpha) \\
0 & 0
\end{array} \right]
\end{equation}
maps the flywheel moments into scalar torques $\bm{u} = \left[\begin{array}{cc} \tau_{fl} & \tau_{fr} \end{array} \right]^T$ about the flywheel axes $\hat{\bm{a}}_l$ and $\hat{\bm{a}}_r$ (expressed in the base frame).
As long as $\bm{C}$ is full column rank, it is possible to control both roll and pitch angles. If $\alpha = 0$, the roll results to be uncontrollable through the \gls{ocs}; otherwise, if $\alpha = \pi/2$, the pitch is uncontrollable. 
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/axes.eps}
	\caption{\small Representation of the flywheels rotation axes, seen from the top of the robot.}
	\label{fig:axes}
\end{figure}
The angle is selected considering the ratio of the eigenvalues along the $x$ and $y$ directions of the ellipsoid of the robot inertia(Fig. \ref{fig:ellipsoid}):
\begin{equation*}
\alpha^{*}=\tan^{-1}\left(\dfrac{I_{r, xx}}{I_{r, yy}}\right).
\end{equation*}
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/inertiaEllipsoid.eps}
	\caption{\small The inertia tensor can be seen as an ellipsoid. Its principal axes in the direction of the eigenvectors of the tensor and their length depends on the eigenvalues.}
	\label{fig:ellipsoid}
\end{figure}
In the case of Solo12, this angle is about $40^{\circ}$. With these considerations, the angular momentum produced by the wheels, expressed in the robot base frame, is:
\begin{equation}
\begin{split}
\tensor[_b]{\bm{L}}{_f} &= \tensor[_b]{\bm{L}}{_{fl}} + \tensor[_b]{\bm{L}}{_{fr}} \\
&=I_{f, zz} \, \bm{\omega}_{fl} + I_{f, zz} \, \bm{\omega}_{fr}\\
&=I_{f, zz}
\left[\begin{array}{c}
\left(\omega_{fl} - \omega_{fr}\right)\sin(\alpha) \\ 
\left(\omega_{fl} +  \omega_{fr}\right)\cos(\alpha) \\
0
\end{array}\right]
\end{split}
\label{eq:bLf}
\end{equation}
in which $\bm{\omega}_{fl}= \omega_{fl} \, \hat{\bm{a}}_{l}$ and $\bm{\omega}_{fr}= \omega_{fr} \, \hat{\bm{a}}_{r}$ are the angular velocity vectors of the two wheels, and $\omega_{fl}$ and $\omega_{fr}$ are the angular speeds provided to each flywheel by its actuation system. The latter equation shows that the difference of the two angular speeds impacts on roll rotations, while their difference can be used to adjust the pitch, see Fig. \ref{fig:sum-diff}. Using the definition of $\bm{C}$, \eqref{eq:bLf} rewrites as
\begin{equation}
\begin{split}
\tensor[_b]{\bm{L}}{_f} & = I_{f,zz}
\left[
\begin{array}{cc}
\sin(\alpha) & -\sin(\alpha) \\
\cos(\alpha) & \cos(\alpha) \\
0 & 0
\end{array}
\right]
\left[
\begin{array}{cc}
\omega_{fl} \\
\omega_{fr}
\end{array}
\right] \\
&= I_{f,zz} \bm{C} \bm{\omega}_f.
\end{split}
\end{equation}
\begin{figure}
	\centering
	\includegraphics[width=.8\linewidth]{figures/sum_diff_mod.eps}
	\caption{\small Having incident rotation axes, the \gls{ocs} allows to control both the robot roll and pitch. Notice that the roll is influenced by the difference of the angular speeds of the flywheels, $\omega_x = \left(\omega_{fl} - \omega_{fr}\right)\sin(\alpha)$. On the contrary, the pitch depends on the sum of the angular speeds $\omega_y = \left(\omega_{fl} -  \omega_{fr}\right)\cos(\alpha)$.}
	\label{fig:sum-diff}
\end{figure}
